### 2025-01-21 article/article.tex
WIP

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\imp}{\Rightarrow}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\inv}{^{-1}}
\newcommand{\eps}{\epsilon}
\newcommand{\type}{\texttt{Type}}
\newcommand{\prop}{\texttt{Prop}}
\newcommand{\prun}{\texttt{pr}_1}
\newcommand{\prde}{\texttt{pr}_2}
\newcommand{\inl}{\texttt{inl}}
\newcommand{\inr}{\texttt{inr}}
\newcommand{\skl}{\vspace{\baselineskip}}
\newcommand{\gui}[1]{\og{}#1\fg{}}
\newcommand{\Xge}[1]{X^{\ge#1}}
\newcommand{\Xle}[1]{X^{\le#1}}
\newcommand{\Xgt}[1]{X^{>#1}}
\newcommand{\Age}[1]{\mathcal{A}^{\ge#1}}
\newcommand{\Agt}[1]{\mathcal{A}^{>#1}}
\newcommand{\Ale}[1]{\mathcal{A}^{\le#1}}
\newcommand{\Xgtn}{X^{>n}}
\newcommand{\Xgtm}{X^{>m}}
\newcommand{\Xgtk}{X^{>k}}
\newcommand{\Xlen}{X^{\le n}}
\newcommand{\Alen}{\mathcal{A}^{\le n}}
\newcommand{\dx}{\mathrm{d}x}
\newcommand{\dy}{\mathrm{d}y}
\newcommand{\prth}[1]{\left(#1\right)}
\renewcommand{\empty}{\varnothing}
\newcommand{\tendl}[1]{\longrightarrow_{#1}}
\newcommand{\dbrack}[1]{\llbracket #1 \rrbracket}
\newcommand{\priv}{\,\backslash\,}
\newcommand{\restr}[1]{\vert_{#1}}
\newcommand{\mub}{\overline{\mu}}
\newcommand{\ox}{\otimes}
\renewcommand{\phi}{\varphi}
\newcommand{\piun}{{\pi_{\dbrack{1,n}}}}
\newcommand{\pizn}{{\pi_{\dbrack{0,n}}}}
\newcommand{\Xint}[1]{X^{\dbrack{#1}}}
\newcommand{\tost}{\to_*}
\newcommand{\ind}[1]{\mathbb{1}_{#1}}
\newcommand{\dmu}{\mathrm{d}\mu}
\newcommand{\mathlib}{\texttt{mathlib}}

@@

	This work was initiated by Degenne. He introduced a compositon of kernels indexed by integer intervals which makes the equivalence invisible.
	\begin{lstlisting}
	def compProdNat {i j k : ℕ} (κ : Kernel ((l : Iic i) → X l)
		((l : Ioc i j) → X l))
		(η : Kernel ((l : Iic j) → X l) ((l : Ioc j k) → X l)) :
		Kernel ((l : Iic i) → X l) ((l : Ioc i k) → X l)
	\end{lstlisting}
	This takes two kernels $\kappa : \Xle{i} \tost X^{\dbrack{i+1,j}}$ and $\eta : \Xle{j} \tost X^{\dbrack{j+1,k}}$ and returns a kernel from $\Xle{i}$ to $X^{\dbrack{i+1,k}}$. Degenne then proves that this composition is associative, and uses it to define
	\begin{lstlisting}
	def kerNat (κ : (k : ℕ) → Kernel ((l : Iic k) → X l) (X (k + 1))) (i j : ℕ) :
		Kernel ((l : Iic i) → X l) ((l : Ioc i j) → X l).
	\end{lstlisting}
	This takes as an argument a family of kernels $(\kappa_k)_{k\in\N}$ with $\kappa_k : \Xle{k}\to X_{k+1}$ just as in Theorem~\ref{thm:it}. Then given two integers $i$ and $j$ it returns a kernel \texttt{kerNat $\kappa$ i j} which is the composition
	$$\kappa_i \ox \kappa_{i+1} \ox \cdots \ox \kappa_{j-1}.$$
	We are therefore now able to speak about those compositions which are necessary to state the theorem. However \texttt{kerNat} proved to be not enough for later purposes. Indeed after proving the theorem and trying to write some API we thought that it would be best to to have a kernel which does not necesarrily starts at time 0 but at any instant. This  would prove nicer for example when it comes to write formulas for the conditional expectation against the Ionescu-Tulcea kernel (see Section). However we bumped into an issue. Having different kernels starting at different times was proving hard to manipulate even with the compositions that were implemented. That is why I chose to add another layer on top of \texttt{kerNat} which proved to work out smoothly.
	\begin{lstlisting}
	def partialKernel (a b : ℕ) :
		Kernel ((i : Iic a) → X i) ((i : Iic b) → X i)
	\end{lstlisting}
	Given integer $a$ and $b$ this function will output a kernel from $\Xle{a}$ to $\Xle{b}$. The idea is that the input of this kernel is the full trajectory up to time $a$ and the output is the distribution of the full trajectory up to time $b$. Of course this distribution is deterministic up to time $a$ and then we compose with \texttt{kerNat $\kappa$ a b}. The advantage of this construction is that in the end we always manipulate kernels with domains and codomains of the form $\Xlen$ for some integer $n$ which makes all the composition much easier. For example if we take $a$, $b$ and $c$ three integers satisfying $a \land c \le b$ we get
	\begin{lstlisting}
	(partialKernel κ b c) ∘ₖ (partialKernel κ a b) = partialKernel κ a c
	\end{lstlisting}
	where $\circ_{\rm k}$ is the composition of kernels. This nice property has the advantage to be expressed with a regular operation on kernels rather than one which would be made up for the occasion. With this tool we get a satisfactory way of expressing the composition of multiple kernels.

@@

	The idea relies on the notion of marginals exposed in Section~4 of [Van2024]. I generalized their idea as follows: given $a$ and $b$ integers, $f : X \to [0,+\infty]$ and $x \in X$ I defined \texttt{lamrginalPartialKernel $\kappa$ a b f x} as
	$$\int_{\Xle{b}} f(y \leadsto_b x) \mathrm{d}\eta_{a,b}(x_0,...,x_a).$$

@@

	This trick of considering some function as defined over $X$ while it really only depends on the coordinates $\ge n$ brings the necessity of introducing a simple way to indicate which variables a given function actually depends on. We therefore introduced the following predicate.
	\begin{lstlisting}
	def DependsOn (f : ((i : ι) → α i) → β) (s : Set ι) : Prop :=
	∀ ⦃x y⦄, (∀ i ∈ s, x i = y i) → f x = f y
	\end{lstlisting}
	The idea is that if $f : \prod_{i\in\iota} \alpha_i \to \beta$ and $s \subset \iota$, \texttt{DependsOn f s} means that to check $f(x) = f(y)$ it is enough to show that for any $i \in s$, $x_i = y_i$. In other words $f$ only depends on coordinates indexed by $s$. For example, if $f$ only depends on coordinates $\le b$ and we integrate over its coordinates between $a$ and $b$ then the output only depends on coordinates $\le a$. This is formalized as follows.
	\begin{lstlisting}
	theorem dependsOn_lmarginalPartialKernel (a : ℕ) {b : ℕ}
		{f : ((n : ℕ) → X n) → ℝ≥0∞}
		(hf : DependsOn f (Iic b)) (mf : Measurable f) :
		DependsOn (lmarginalPartialKernel κ a b f) (Iic a)
	\end{lstlisting}
\end{document}

>>>>

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\imp}{\Rightarrow}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\inv}{^{-1}}
\newcommand{\eps}{\epsilon}
\newcommand{\type}{\texttt{Type}}
\newcommand{\prop}{\texttt{Prop}}
\newcommand{\prun}{\texttt{pr}_1}
\newcommand{\prde}{\texttt{pr}_2}
\newcommand{\inl}{\texttt{inl}}
\newcommand{\inr}{\texttt{inr}}
\newcommand{\skl}{\vspace{\baselineskip}}
\newcommand{\gui}[1]{\og{}#1\fg{}}
\newcommand{\Xge}[1]{X^{\ge#1}}
\newcommand{\Xle}[1]{X^{\le#1}}
\newcommand{\Xgt}[1]{X^{>#1}}
\newcommand{\Age}[1]{\mathcal{A}^{\ge#1}}
\newcommand{\Agt}[1]{\mathcal{A}^{>#1}}
\newcommand{\Ale}[1]{\mathcal{A}^{\le#1}}
\newcommand{\Xgtn}{X^{>n}}
\newcommand{\Xgtm}{X^{>m}}
\newcommand{\Xgtk}{X^{>k}}
\newcommand{\Xlen}{X^{\le n}}
\newcommand{\Alen}{\mathcal{A}^{\le n}}
\newcommand{\dx}{\mathrm{d}x}
\newcommand{\dy}{\mathrm{d}y}
\newcommand{\prth}[1]{\left(#1\right)}
\renewcommand{\empty}{\varnothing}
\newcommand{\tendl}[1]{\longrightarrow_{#1}}
\newcommand{\dbrack}[1]{\llbracket #1 \rrbracket}
\newcommand{\priv}{\,\backslash\,}
\newcommand{\restr}[1]{\vert_{#1}}
\newcommand{\mub}{\overline{\mu}}
\newcommand{\ox}{\otimes}
\renewcommand{\phi}{\varphi}
\newcommand{\piun}{{\pi_{\dbrack{1,n}}}}
\newcommand{\pizn}{{\pi_{\dbrack{0,n}}}}
\newcommand{\Xint}[1]{X^{\dbrack{#1}}}
\newcommand{\tost}{\to_*}
\newcommand{\ind}[1]{\mathbb{1}_{#1}}
\newcommand{\dmu}{\mathrm{d}\mu}
\newcommand{\dlamb}{\mathrm{d}\lambda}
\newcommand{\mathlib}{\texttt{mathlib}}

@@

	This work was initiated by Degenne. He introduced a compositon of kernels indexed by integer intervals which makes the equivalence invisible.
	\begin{lstlisting}
	def compProdNat {i j k : ℕ} (κ : Kernel ((l : Iic i) → X l)
		((l : Ioc i j) → X l))
		(η : Kernel ((l : Iic j) → X l) ((l : Ioc j k) → X l)) :
		Kernel ((l : Iic i) → X l) ((l : Ioc i k) → X l)
	\end{lstlisting}
	This takes two kernels $\kappa : \Xle{i} \tost X^{\dbrack{i+1,j}}$ and $\eta : \Xle{j} \tost X^{\dbrack{j+1,k}}$ and returns a kernel from $\Xle{i}$ to $X^{\dbrack{i+1,k}}$. Degenne then proves that this composition is associative, and uses it to define
	\begin{lstlisting}
	def kerNat (κ : (k : ℕ) → Kernel ((l : Iic k) → X l) (X (k + 1))) (i j : ℕ) :
		Kernel ((l : Iic i) → X l) ((l : Ioc i j) → X l).
	\end{lstlisting}
	This takes as an argument a family of kernels $(\kappa_k)_{k\in\N}$ with $\kappa_k : \Xle{k}\to X_{k+1}$ just as in Theorem~\ref{thm:it}. Then given two integers $i$ and $j$ it returns a kernel \lean{kerNat κ i j} which is the composition
	$$\kappa_i \ox \kappa_{i+1} \ox \cdots \ox \kappa_{j-1}.$$
	We are therefore now able to speak about those compositions which are necessary to state the theorem. However \lean{kerNat} proved to be not enough for later purposes. Indeed after proving the theorem and trying to write some API we thought that it would be best to to have a kernel which does not necesarrily starts at time 0 but at any instant. This  would prove nicer for example when it comes to write formulas for the conditional expectation against the Ionescu-Tulcea kernel (see Section). However we bumped into an issue. Having different kernels starting at different times was proving hard to manipulate even with the compositions that were implemented. That is why I chose to add another layer on top of \lean{kerNat} which proved to work out smoothly.
	\begin{lstlisting}
	def partialKernel (a b : ℕ) :
		Kernel ((i : Iic a) → X i) ((i : Iic b) → X i)
	\end{lstlisting}
	Given integer $a$ and $b$ this function will output a kernel from $\Xle{a}$ to $\Xle{b}$. The idea is that the input of this kernel is the full trajectory up to time $a$ and the output is the distribution of the full trajectory up to time $b$. Of course this distribution is deterministic up to time $a$ and then we compose with \lean{kerNat κ a b}. The advantage of this construction is that in the end we always manipulate kernels with domains and codomains of the form $\Xlen$ for some integer $n$ which makes all the composition much easier. For example if we take $a$, $b$ and $c$ three integers satisfying $a \land c \le b$ we get
	\begin{lstlisting}
	(partialKernel κ b c) ∘ₖ (partialKernel κ a b) = partialKernel κ a c
	\end{lstlisting}
	where $\circ_{\rm k}$ is the composition of kernels. This nice property has the advantage to be expressed with a regular operation on kernels rather than one which would be made up for the occasion. With this tool we get a satisfactory way of expressing the composition of multiple kernels.

@@

	The idea relies on the notion of marginals exposed in Section~4 of [Van2024]. I generalized their idea as follows: given $a$ and $b$ integers, $f : X \to [0,+\infty]$ and $x \in X$ I defined \lean{lamrginalPartialKernel κ a b f x} as
	$$\int_{\Xle{b}} f(y \leadsto_b x) \mathrm{d}\eta_{a,b}(x_0,...,x_a).$$

@@

	This trick of considering some function as defined over $X$ while it really only depends on the coordinates $\ge n$ brings the necessity of introducing a simple way to indicate which variables a given function actually depends on. We therefore introduced the following predicate.
	\begin{lstlisting}
	def DependsOn (f : ((i : ι) → α i) → β) (s : Set ι) : Prop :=
	∀ ⦃x y⦄, (∀ i ∈ s, x i = y i) → f x = f y
	\end{lstlisting}
	The idea is that if $f : \prod_{i\in\iota} \alpha_i \to \beta$ and $s \subset \iota$, \lean{DependsOn f s} means that to check $f(x) = f(y)$ it is enough to show that for any $i \in s$, $x_i = y_i$. In other words $f$ only depends on coordinates indexed by $s$. For example, if $f$ only depends on coordinates $\le b$ and we integrate over its coordinates between $a$ and $b$ then the output only depends on coordinates $\le a$. This is formalized as follows.
	\begin{lstlisting}
	theorem dependsOn_lmarginalPartialKernel (a : ℕ) {b : ℕ}
		{f : ((n : ℕ) → X n) → ℝ≥0∞}
		(hf : DependsOn f (Iic b)) (mf : Measurable f) :
		DependsOn (lmarginalPartialKernel κ a b f) (Iic a)
	\end{lstlisting}

	\section{Writing an API}

	When formalizing a new mathematical object one usually does not stop at writing the definition. Indeed the objected is intended to be usable by all the users without having to know how the object was implemented. That is why it is important to accompany the object with many results which allow other users to use the object without knowing the implementation details.

	We now present some of these results. The API is still very basic and will be developed as other users pick up the object to formalize more objects. The best way to find new lemmas to formalize is to choose a target statement and try and formalize it. In our case we aim to prove the following result.

	\begin{nota}
		For $n \in \N$ we set $\F_n := \pi_{\dbrack{0,n}}\inv\prth{\Alen}$.
	\end{nota}

	\begin{thm}
		\label{thm:condexp}
		Let $a, b \in \N$ with $a \le b$. Let $u \in \Xle{a}$. Let $f : X \to \R$ be an integrable function with respect to $\lambda_a(u)$. Then for $\lambda_a(u)$-almost every $x := (x_n)_{n\in\N}$, $f$ is integrable with respect to $\lambda_b(x_0,...,x_b)$ and
		$$\E_{\lambda_a(u)}[f \mid \F_n] (x) = \int_X f(y) \lambda_b(x_0,...,x_b, \dy).$$
	\end{thm}

	Let us give the proof to highlight the necessary intermediate results. Take $A \in \Ale{b}$ and set $\pi := \pi_{\dbrack{0,b}}$. Our goal is to prove that
	$$\int_{\pi\inv(A)} \int_X f(y) \lambda_b(\pi(x), \dy)\lambda_a(u,\dx) = \int_{\pi\inv(A)} f \dlamb_a.$$
	We start by rewriting the outer integral using the push-forward measure:
	$$\int_{\pi\inv(A)} \int_X f(y) \lambda_b(\pi(x), \dy)\lambda_a(u,\dx) = \int_A \int_X f(y) \lambda_b(\pi(x), \dy)(\pi_*\lambda_a)(u,\dx).$$
	The first we will need is that $\pi_*\lambda_a = \eta_{a,b}$. This is quite obvious: as $\lambda_a$ gives the whole trajectory, keeping only the cooordinates up to time $b$ yields the trajectory up to time $b$. This is formalized as below.
	\begin{lstlisting}
theorem ionescuTulceaKernel_proj (a b : ℕ) :
(ionescuTulceaKernel κ a).map (frestrictLe b) = partialKernel κ a b
	\end{lstlisting}
	So we get $$\int_{\pi\inv(A)} \int_X f(y) \lambda_b(\pi(x), \dy)\lambda_a(u,\dx) = \int_A \int_X f(y) \lambda_b(\pi(x), \dy)\eta_{a,b}(u,\dx).$$
\end{document}
